---
title: "Building a Gibbs Sampler with dbarts"
author: "Vincent Dorie"
date: "2026-08-19"
output: pdf_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Building a Gibbs Sampler with dbarts}
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For users who are comfortable with writing their own posterior samplers, `dbarts` makes it easy to incorporate a linear BART component in a Gibbs sampler. In the example below, we fit a functional mixture model, where observations may have come from one of two underlying functions and class membership is unobserved.

## Model

Performing inference conditional on $x$, we assume that:

$$Y\mid Z = z \sim \mathrm{N}\left(f(x, z), \sigma^2 \right),$$
$$Z \sim \mathrm{Bern}(p),$$

and that $Z$ is unobserved. To make the model fully Bayesian, we impose a $\mathrm{Beta}(1, 1)$ prior on $p$. $f$ and $\sigma^2$ have implicit priors defined by BART.

## Conditional Posteriors

To implement a Gibbs sampler, we will have to produce updates for $Z$ and $p$. After writing down the joint distribution of $Z$, $Y$, $p$, $f$, and $\sigma^2$, we have:

* $P(Z = 1 \mid Y = y, p, f, \sigma^2) \propto \phi(y, f(x, 1), \sigma) \cdot p$
* $P(Z = 0 \mid Y = y, p, f, \sigma^2) \propto \phi(y, f(x, 0), \sigma) \cdot (1 - p)$
* $p \mid Z = z\sim \mathrm{Beta}(1 + \sum z_i, 1 + \sum (1 - z_i))$

where $\phi$ is the density of a normal distribution. This yields an overall strategy of:

1. Draw a sample from BART
   1. Draw samples of $f(x, 1)$ and $f(x, 0)$
   2. Draw a sample of $\sigma^2$
2. Use $f(x, 1)$, $f(x, 0)$, and $p$ to draw a sample of $Z$
3. Use $Z$ to draw a sample of $p$

## Simulated Data

To illustrate, we create some toy data. The simulated data is visualized later, together with the fitted model.

```{r generate_data}
# Underlying functions
f0 <- function(x) 90 + exp(0.06 * x)
f1 <- function(x) 72 + 3 * sqrt(x)

set.seed(2793)

# Generate true values.
n <- 120
p.0 <- 0.5

z.0 <- rbinom(n, 1, p.0)
n1 <- sum(z.0); n0 <- n - n1

# In order to make the problem more interesting, x is confounded with both
# y and z.
x <- numeric(n)
x[z.0 == 0] <- rnorm(n0, 20, 10)
x[z.0 == 1] <- rnorm(n1, 40, 10)
y <- numeric(n)
y[z.0 == 0] <- f0(x[z.0 == 0])
y[z.0 == 1] <- f1(x[z.0 == 1])
y <- y + rnorm(n)

data_train <- data.frame(y = y, x = x, z =     z.0)
data_test  <- data.frame(       x = x, z = 1 - z.0)
```

## Implementing a Gibbs Sampler

For this specific model, the sampler needs to be updated with new predictor variables/covariates. This can be done with a `dbartsSampler` reference class object by calling the `sampler$setPredictor` function. As, given any $Z = z$, the sampler also needs to evaluate $f(x, 1 - z)$, we utilize the test slot of the sampler and keep it up-to-date with the "counterfactual" predictor using the `sampler$setTestPredictor` function. Models that instead modify the response variable can use the `sampler$setResponse` or `sampler$setOffset` functions.

One complication in updating the predictor matrix while the sampler runs is that new values of $z$  must leave the sampler in an internally consistent state. During warmup this constraint is ignored; after warmup is complete, rejection sampling is used to guarantee that no leaf nodes are empty. This is accomplished by using the `forceUpdate` argument to `setPredictor` and checking that the logical response is `TRUE`.

```{r gibbs_sampler}
n_warmup  <- 100
n_samples <- 500
n_total   <- n_warmup + n_samples

# Allocate storage for result.
samples_p   <- rep.int(NA_real_, n_samples)
samples_z   <- matrix(NA_real_, n_samples, n)
samples_mu0 <- matrix(NA_real_, n_samples, n)
samples_mu1 <- matrix(NA_real_, n_samples, n)

library(dbarts, quietly = TRUE)

# We only need to draw one sample at a time, although for illustrative purposes
# a small degree of thinning is done to the BART component.
control <- dbartsControl(updateState = FALSE, verbose = FALSE,
                         n.burn = 0L, n.samples = 1L, n.thin = 3L,
                         n.chains = 1L)
# We create the sampler with a z vector that contains at least one 1 and one 0,
# so that all of the cut points are set correctly.
sampler <- dbarts(y ~ x + z, data_train, data_test, control = control)

# Sample from prior.
p <- rbeta(1, 1, 1)
z <- rbinom(n, 1, p)

# Ignore result of this sampler call
invisible(sampler$setPredictor(x = z, column = 2, forceUpdate = TRUE))
sampler$setTestPredictor(x = 1 - z, column = 2)
sampler$sampleTreesFromPrior()

for (i in seq_len(n_total)) {
  # Draw a single sample from the posterior of f and sigma^2.
  samples <- sampler$run()

  # Recover f(x, 1) and f(x, 0).
  mu0 <- ifelse(z == 0, samples$train[,1], samples$test[,1])
  mu1 <- ifelse(z == 1, samples$train[,1], samples$test[,1])
  
  p0 <- dnorm(y, mu0, samples$sigma[1]) * (1 - p)
  p1 <- dnorm(y, mu1, samples$sigma[1]) * p
  p.z <- p1 / (p0 + p1)
  z <- rbinom(n, 1, p.z)
  
  if (i <= n_warmup) {
    sampler$setPredictor(x = z, column = 2, forceUpdate = TRUE)
  } else while (sampler$setPredictor(x = z, column = 2) == FALSE) {
    z <- rbinom(n, 1, p.z)
  }
  sampler$setTestPredictor(x = 1 - z, column = 2)
  
  n1 <- sum(z); n0 <- n - n1
  p <- rbeta(1, 1 + n1, 1 + n0)

  # Store samples if no longer warming up.
  if (i > n_warmup) {
    offset <- i - n_warmup
    samples_p[offset] <- p
    samples_z[offset,] <- z
    samples_mu0[offset,] <- mu0
    samples_mu1[offset,] <- mu1
  }
}
```

## `Save`ing the sampler

To use `save` and `load` on the sampler, it must first be told to write its state out as an `R` object:

```{r sampler_state}
sampler$storeState()
```

## Visualizing the Result

Finally, we visualize the results. In the graph below, the estimated label of observations encircles the true labels, both of which are represented by point color. We see that, beyond the range of the observed data, the estimated functions regress towards each other and points start to be mislabeled.

```{r result_visualization}
mean_mu0 <- apply(samples_mu0, 2, mean)
mean_mu1 <- apply(samples_mu1, 2, mean)

ub_mu0 <- apply(samples_mu0, 2, quantile, 0.975)
lb_mu0 <- apply(samples_mu0, 2, quantile, 0.025)

ub_mu1 <- apply(samples_mu1, 2, quantile, 0.975)
lb_mu1 <- apply(samples_mu1, 2, quantile, 0.025)

curve(f0(x), 0, 80, ylim = c(80, 110), ylab = expression(f(x)),
      main = "Mixture Model")
curve(f1(x), add = TRUE)

points(x, y, pch = 20, col = ifelse(z.0 == 0, "black", "gray"))

lines(sort(x), mean_mu0[order(x)], col = "red")
lines(sort(x), mean_mu1[order(x)], col = "red")

# Add point-wise confidence intervals.
lines(sort(x), ub_mu0[order(x)], col = "gray", lty = 2)
lines(sort(x), lb_mu0[order(x)], col = "gray", lty = 2)

lines(sort(x), ub_mu1[order(x)], col = "gray", lty = 3)
lines(sort(x), lb_mu1[order(x)], col = "gray", lty = 3)

# Without constraining z for an observation to 0/1, its interpretation may be
# flipped from that which generated the data.
mean_z <- ifelse(apply(samples_z, 2, mean) <= 0.5, 0L, 1L)
if (mean(mean_z != z.0) > 0.5) mean_z <- 1 - mean_z

points(x, y, pch = 1, col = ifelse(mean_z == 0, "black", "gray"))
legend("topright", c("true func", "est func", "group 0", "group 1",
                     "est group 0", "est group 1"),
       lty = c(1, 1, NA, NA, NA, NA), pch = c(NA, NA, 20, 20, 1, 1),
       col = c("black", "red", "black", "gray", "black", "gray"),
       cex = 0.8, box.col = "white", bg = "white")
```

## Several Forests with Multipliers

The sampler above holds one function. A model may want several, each entering
the response through a multiplier the outer loop knows:

$$Y_i \sim \mathrm{N}\left(\sum_{f=1}^{K} m_{f,i}\, g_f(x_i),\ \sigma^2\right),$$

where $g_1, \ldots, g_K$ are separate sums of trees and $m_{f,i}$ is a known
per-observation scalar - an exposure, a dose, a treatment indicator, a group
loading. Bayesian causal forests is the $K = 2$ case with $m_1 \equiv 1$ and
$m_2 = z$.

One `dbartsSampler` per forest composes this in R, and the composition is
ordinary backfitting once the multiplier is folded in. Given the residual
$r_i = y_i - \sum_{g \neq f} m_{g,i} g_g(x_i)$, forest $f$'s leaf conditionals
want response $r_i / m_{f,i}$ and precision weight $w_i m_{f,i}^2$; those are
the two channels the loop writes, with `setSigma` for the third and
`run(0L, 1L)` to advance one sweep. `updateScale = FALSE` (the default) is
what makes the swap coherent: every sweep's residual is read on the scale
creation fixed, so the leaf prior keeps meaning what it meant.

A multiplier at or near zero says the row carries no information about that
forest. Divide anyway and the response is amplified without bound, so the
recipe snaps such a row to an exactly zero response paired with its exactly
zero weight - which is what the engine does internally, at
$\sqrt{\texttt{.Machine\$double.eps}}$.

```{r kforest}
set.seed(919)
n.mf <- 200L
x.mf <- matrix(runif(n.mf * 3L), n.mf, 3L)
z.mf <- rbinom(n.mf, 1L, 0.5)
g1.true <- 2 * sin(pi * x.mf[, 1L])
g2.true <- 1 + 2 * x.mf[, 2L]
y.mf <- g1.true + z.mf * g2.true + rnorm(n.mf, 0, 0.5)

# the known per-observation multipliers, one vector per forest
m <- list(rep(1, n.mf), as.double(z.mf))
K <- length(m)

control.mf <- dbartsControl(n.chains = 1L, n.threads = 1L, n.trees = 25L,
                            updateState = FALSE)
# every sampler is built on the SAME response, with k fixed and sigma owned
# by the outer loop
samplers <- lapply(seq_len(K), function(f)
    dbarts(x.mf, y.mf, control = control.mf, node.prior = normal(k = 1),
           resid.prior = fixed(1)))
# one prior budget divided among the forests, rather than K copies of the
# budget for the whole of the sum
base.mf <- samplers[[1L]]$getCalibration()[1L, "prior.scale"]
for (s in samplers) s$setCalibration(prior.scale = base.mf / sqrt(K))

g <- rep(list(rep(0, n.mf)), K)
sigma.mf <- 1
n.keep <- 200L
keep.g2 <- matrix(0, n.mf, n.keep)

for (i in seq_len(100L + n.keep)) {
  for (f in seq_len(K)) {
    r <- y.mf - Reduce(`+`, Map(`*`, m[-f], g[-f]))
    zero <- abs(m[[f]]) < sqrt(.Machine$double.eps)
    samplers[[f]]$setResponse(ifelse(zero, 0, r / ifelse(zero, 1, m[[f]])),
                              updateScale = FALSE)
    samplers[[f]]$setWeights(m[[f]]^2)
    samplers[[f]]$setSigma(sigma.mf)
    samplers[[f]]$run(0L, 1L)
    g[[f]] <- as.numeric(samplers[[f]]$getFitsWithoutOffset())
  }
  # sigma is drawn in R, against the COMBINED residual
  resid.mf <- y.mf - Reduce(`+`, Map(`*`, m, g))
  sigma.mf <- sqrt(sum(resid.mf^2) / rchisq(1L, n.mf))
  if (i > 100L) keep.g2[, i - 100L] <- g[[2L]]
}

c(recovery = cor(rowMeans(keep.g2), g2.true), sigma = sigma.mf, truth = 0.5)
```

### What it costs

Measured on one machine against a single batched engine sampler at matched
total trees - the denominator matters, and a price quoted without one says
nothing - K samplers driven a sweep at a time cost about 1.9-2.1 times that
sampler at $K = 2$ and 4.6-5.2 times at $K = 8$, growing roughly linearly in
$K$: each sampler pays its own per-sweep response install, weight install and
result allocation over the full data. Driving one sampler a sweep at a time
rather than in a batch is a separate and much smaller charge, about 1.1x.

### Five things the recipe does not get

* **No joint move on the multipliers.** If the outer loop also draws the
  $m_f$, it draws them in a block alternating with the forests. A sampler that
  holds both can interweave them instead, and the two mix differently whenever
  a forest and its multiplier are jointly weakly identified.
* **The division is yours.** The caller forms $r / m_f$ and so owns the
  conditioning, including the snap above.
* **$K$ quantized predictor stores and $K$ sampler states**, where one
  sampler holding $K$ forests keeps one of each.
* **$K$ leaf calibrations to state**, one per sampler, rather than a single
  calibration stated across the forests at creation.
* **A serial outer loop.** The forests are drawn one after another in R, so
  the composition cannot spread them across threads.

### Graduating to a multi-forest sampler

`dbarts(..., forests = list(forest(), forest(basis = ~ factor(z))))` fits this
model in one sampler, and the question a prototype should be able to ask is
whether moving to it changes the posterior being targeted.

The additive decomposition ports exactly. What does not port by itself is
prior shape. The two routes agree only if the caller

1. centres $y$ at its midrange, or absorbs the response transform's shift with
   `setOffset`: each single-forest sampler's reported fit carries that shift,
   so the composition attributes it once per forest where the sampler
   attributes it once;
2. pins the leaf prior identically - `node.prior = normal(k)` with $k$ fixed,
   every sampler built on a common $y$, and `updateScale = FALSE` throughout;
3. accepts that with a multiplier far from 1 the internal response a forest
   sees leaves the band its leaf prior was calibrated for.

Without (1) and (2) the two routes are different priors that agree closely.
With them the remaining difference is the leaf calibration alone. Measured on
a pinned two-forest sampler against the recipe above, at $n = 400$ over eight
chains: with the remedies the posterior mean conditional effect agreed to
3.4% pointwise, inside Monte Carlo error in root-mean-square; without them the
difference was 11.9% pointwise and several times Monte Carlo error. Both
recovered the truth about equally well, which is the point - the difference is
a prior, not a defect.

State equally plainly what the recipe does not reach and never will: a
response family, link or latent structure the sampler does not provide, and a
leaf or hyperprior law it does not provide. Those are selected by argument,
not composed in R. `forest()` carries the same scale knobs the composition
sets by hand, and `$setForestBasis`, `$setForestWeights` and
`$getForestAmplitudes` are the post-creation routes into a multi-forest
sampler, so a prototype ports without changing the model it targets.

## Response Families

Everything above fits a continuous response, `dbarts`'s default (`family = "auto"`, which chooses `"gaussian"` for a continuous $y$ and `"probit"` for one coded 0/1). `family` can also be set to `"gaussian"`, `"probit"`, or `"logistic"` explicitly. The two binary families model $P(Y = 1 \mid x) = F(f(x))$ for a link $F$ and the sum-of-trees $f$, and report fits and predictions on that latent scale rather than as probabilities (`pnorm`/`plogis` recovers the probability from it). Probit takes $F$ to be the standard normal CDF; logistic uses Polya-Gamma augmentation and takes $F$ to be the logistic CDF, whose latent variable has heavier tails than probit's, so the leaf-value prior widens to match (`node.scale` becomes `pi * sqrt(3)` in place of probit's `3`, keeping the same span of $f(x)$ relative to the response).

```{r family_example}
set.seed(11)
n.fam <- 60
x.fam <- runif(n.fam)
y.fam <- rbinom(n.fam, 1, plogis(3 * (x.fam - 0.5)))

fit.logistic <- dbarts(y.fam ~ x.fam, family = "logistic",
                       control = dbartsControl(n.trees = 10L, n.chains = 1L,
                                                n.threads = 1L))
samples.logistic <- fit.logistic$run(20L, 20L)
```

## DART

When only a handful of predictors out of many are relevant, the CGM tree prior's uniform choice of split variable wastes proposals on the rest. `tree.prior = dart(...)` replaces it with a Dirichlet prior over split-variable probabilities (Linero 2018): the sampler learns which variables matter and increasingly proposes splits on them, which doubles as a variable-importance diagnostic.

```{r dart_example}
set.seed(12)
n.dart <- 60
x.dart <- matrix(runif(n.dart * 8), n.dart)
y.dart <- 4 * x.dart[, 1] + rnorm(n.dart, 0, 0.2)

fit.dart <- dbarts(y.dart ~ x.dart, tree.prior = dart(),
                   control = dbartsControl(n.trees = 10L, n.burn = 100L,
                                            n.chains = 1L, n.threads = 1L))
samples.dart <- fit.dart$run(100L, 30L)

# One row per predictor, averaged over the retained samples: only the first
# column of x.dart drives y.dart, and its split probability dominates.
round(rowMeans(samples.dart$varprobs), 3)
```

The Dirichlet concentration `alpha` is itself sampled by default (`update.alpha = TRUE`); a smaller `alpha` concentrates the split probabilities more aggressively. `?dbartsPriors` documents the rest of `dart`'s arguments.

## Data Frames, Factors, and Missing Values

`dbarts` and `bart2` take a `data.frame` directly, and (`factors = "categorical"`, the default for both) an unordered factor predictor becomes a single categorical column: its splits send a subset of levels down each branch, rather than the one-indicator-per-level expansion `bart` still uses. `factors = "indicators"` recovers that expansion on either function, useful for comparing the two. Ordered factors are their own predictor kind regardless of `factors`: coded on their integer levels and split at a threshold, with one candidate threshold per level boundary rather than a fixed count of cuts, so `n.cuts` does not reach them.

Missing predictor values need not be removed beforehand. The default `missing = "incorporate"` gives every split rule a learned direction for `NA`: an observation whose split variable is missing follows that rule's direction ("Missing Incorporated in Attributes", Twala et al. 2008), estimated jointly with the rest of the tree. `missing = "error"` instead rejects any predictor containing `NA`.

```{r missing_example}
set.seed(13)
n.miss <- 60
g <- factor(sample(c("a", "b", "c"), n.miss, replace = TRUE))
x.miss <- rnorm(n.miss)
x.miss[sample(n.miss, 10)] <- NA
y.miss <- ifelse(g == "a", 2, -1) +
    ifelse(is.na(x.miss), 0, x.miss) + rnorm(n.miss, 0, 0.2)

df.miss <- data.frame(y = y.miss, x = x.miss, g = g)
fit.miss <- bart2(y ~ x + g, df.miss, n.trees = 10L, n.chains = 1L,
                  n.threads = 1L, n.samples = 20L, n.burn = 20L,
                  verbose = FALSE)
```

`g` above ingests as one categorical predictor with three levels rather than two indicator columns, and the ten missing values of `x` are incorporated rather than dropped.

## Multiple Threading

When implementing a Gibbs sampler using `dbarts` in R, it is most often the case that multiple threads will need to be handled by creating separate copies of the sampler and data. While `dbartsSampler`s are natively multithreaded, few of their slots are stored independently across chains. For an example of this approach and a more complete implementation of the principles in this document, consult the implementation of [rbart_vi](https://github.com/vdorie/dbarts/blob/master/R/rbart.R).

This custom-loop bookkeeping is only needed when embedding BART in a larger sampler. A plain multi-chain `bart2`/`rbart_vi` fit already tracks its chain dimension: `summary()` reports split-R-hat and effective sample size for the scalar parameters (`sigma`, `k`, and, for `rbart_vi`, `tau`) when the `posterior` package is installed, and `as_draws_array`/`as_draws_df` convert its chain-dimensioned draws to `posterior`'s array/data frame conventions.