---
title: "Get Prior"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

## Objective

The goal of this vignette is to show you how to get the prior distribution of the treatment effect.

## Data Generating Process

$$
\begin{align*}
Z &\sim \mbox{Bernoulli}(0.5) \\
X | Z = 1 &\sim N(40,10^2) \\
X | Z = 0 &\sim N(20,10^2) \\
Y(0) | X &\sim N(72 + 3\sqrt{X},1) \\
Y(1) | X &\sim N(90 + exp(0.06X),1) 
\end{align*}
$$

### Plot the DGP

```{r dgp, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
library(ggplot2)
library(dplyr)
library(dbarts)
library(furrr)
plan(multiprocess)

f_treated <- function(x) 90 + exp(0.06*x)
f_control <- function(x) 72 + 3*sqrt(x)

p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) + 
  stat_function(fun = f_control, aes(color = "truth - control")) +
  stat_function(fun = f_treated, aes(color = "truth - treatment")) +
  scale_color_manual("", values = c("red", "blue")) + 
  theme(legend.position="bottom") + 
  xlim(6,62)

p  
```

## Synthetic data

```{r fake_data, warning=FALSE, message=FALSE}
set.seed(123)
N <- 120

my_data <-
  tibble(Z = sample(
    x = c("treatment", "control"),
    size = N,
    replace = TRUE,
    p = c(0.5, 0.5)
  )) %>%
  rowwise()  %>%
  mutate(
    treatment = case_when(Z == "treatment" ~ 1,
                          Z == "control" ~ 0),
    X = case_when(
      Z == "treatment" ~ rnorm(1, mean = 40, sd = 10),
      Z == "control" ~ rnorm(1, mean = 20, sd = 10)
    ),
    Y1 = rnorm(1, mean = 90 + exp(0.06 * X), sd = 1),
    Y0 = rnorm(1, mean = 72 + 3 * sqrt(X), sd = 1),
    Y = case_when(Z == "treatment" ~ Y1,
                  Z == "control" ~ Y0),
    tau = Y1 - Y0
  ) %>%
  tidyr::drop_na() #  remove NAs from negative X's

```

The true Sample Average Treatment Effect in these syntetic data is `r round(mean(my_data$tau)/sd(my_data$Y),2)` standard deviations.



```{r plot_samples, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
ggplot(data = my_data, aes(x=X, y=Y, color = Z)) + 
   geom_point() + 
   stat_function(fun = f_control, aes(color = "truth - control")) +
   stat_function(fun = f_treated, aes(color = "truth - treatment")) + 
   scale_color_manual("", values = c("red", "blue","red", "blue")) + 
   theme(legend.position="bottom") + 
   xlim(6,62)
```



## Fit the model

```{r fit}
train_data <- dbartsData(my_data %>% select(X, treatment), my_data$Y)

fit <- bart2(
  train_data,
  keepTrees       = T,
  verbose         = F,
  n.thread        = 6,
  combineChains   = TRUE
)
## Get the posterior distribution for the treatment effect

x0 <- my_data %>% mutate(treatment = 0) %>% select(X, treatment)
x1 <- my_data %>% mutate(treatment = 1) %>% select(X, treatment)

pred0 <- predict(fit, as.matrix(x0))

pred1 <- predict(fit, as.matrix(x1))

tau_posterior_draws <- pred1 - pred0

SATE_posterior <- rowMeans(tau_posterior_draws)


```


### Check Sigma

```{r check_sigma, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
ggplot(data = tibble(sigma = fit$sigma, draw = 1:length(fit$sigma)),
       aes(y = sigma, x = draw)) +
  geom_line()
neff_sigma <- coda::effectiveSize(x = fit$sigma)

```

Sigma's effective sample size is `r neff_sigma`.

### How well did we recover the DGP?

```{r plot_bart, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
fake_data <-
  expand.grid(X = seq(min(my_data$X), max(my_data$X), length = 1000),
              treatment = c(1, 0)) %>%
  as_tibble() %>%
  mutate(trueY = ifelse(treatment == 1, 90 + exp(0.06 * X), 72 + 3 * sqrt(X)))

predictions <- predict(fit, fake_data)

fake_data <- fake_data %>% 
  mutate(bart_pred = colMeans(predictions))

# add annotations to plot
xstart <- 20-15
length <- 4
text_left <- 26-15
yVals <- seq(150,180,by=4)

ggplot(data = my_data, aes(x = X, y = Y, color = as.factor(treatment))) +
  theme_bw() +
  geom_point() +
  stat_function(fun = f_control, aes(color = "truth - control")) +
  stat_function(fun = f_treated, aes(color = "truth - treatment")) +
  geom_line(data = fake_data,
            aes(x = X, y = bart_pred, color = as.factor(treatment)),
            linetype = "dashed") +
  theme(legend.position = "none") +
  scale_color_manual("",
                     values = c("red", "blue", "red", "blue", "red", "blue", "red", "blue")) +
  annotate(
    geom = "segment",
    x = xstart,
    y = yVals[1],
    xend = xstart + length,
    yend = yVals[1],
    color = "red"
  ) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[1],
    label = c("truth - control"),
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = xstart,
    y = yVals[2],
    xend = xstart + length,
    yend = yVals[2],
    color = "blue"
  ) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[2],
    label = c("truth - treatment"),
    hjust = 0
  ) +
  geom_point(aes(x = xstart + length / 2, y = yVals[3]), color = c("red")) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[3],
    label = c("simulated data - control"),
    hjust = 0
  ) +
  geom_point(aes(x = xstart + length / 2, y = yVals[4]), color = c("blue")) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[4],
    label = c("simulated data - treatment"),
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = xstart,
    y = yVals[5],
    xend = xstart + length,
    yend = yVals[5],
    color = "red",
    linetype = "dashed"
  ) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[5],
    label = c("BART - control"),
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = xstart,
    y = yVals[6],
    xend = xstart + length,
    yend = yVals[6],
    color = "blue",
    linetype = "dashed"
  ) +
  annotate(
    geom = "text",
    x = text_left,
    y = yVals[6],
    label = c("BART - treatment"),
    hjust = 0
  ) 
```


### Plot Posterior

```{r plot_posterior, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
ggplot(data = tibble(value = SATE_posterior / sd(my_data$Y)), aes(x = value)) +
  geom_histogram() +
  geom_vline(xintercept = mean(my_data$tau) / sd(my_data$Y),
             linetype = "dashed") +
  ggtitle("Posterior Distribution of the Sample Average Treatment Effect") +
  xlab("Standard Deviations")
```


## Get the prior distribution for the treatment effect

```{r prior, fig.height=5, fig.width=7, message=FALSE, fig.align='center'}
.getPrior <- function(i, sampler_x0, sampler_x1) {
  sampler_x0$sampleTreesFromPrior()
  sampler_x0$sampleNodeParametersFromPrior()
  prior_samples_x0 <- sampler_x0$predict(x0)
  sampler_x1$sampleTreesFromPrior()
  sampler_x1$sampleNodeParametersFromPrior()
  prior_samples_x1 <- sampler_x1$predict(x1)
  prior_samples_tau <- prior_samples_x1 - prior_samples_x0
  prior_tau_i <- mean(prior_samples_tau)
  return(prior_tau_i)
}

getTreatmentPrior <-
  function(dbarts_data,
           treatment,
           n_samples = 10000,
           parallel = FALSE) {
    sampler_x0 <-
      dbarts(train_data, control = dbartsControl(n.chains = 1))
    sampler_x0$setPredictor(rep_len(0, nrow(my_data)), "treatment")
    
    sampler_x1 <-
      dbarts(train_data, control = dbartsControl(n.chains = 1))
    sampler_x0$setPredictor(rep_len(1, nrow(my_data)), "treatment")
    
    prior_samples_x0 <- matrix(NA, nrow(my_data), n_samples)
    prior_samples_x1 <- matrix(NA, nrow(my_data), n_samples)
    prior_samples_tau <- matrix(NA, nrow(my_data), n_samples)
    
    if (parallel) {
      SATE_prior <- future_map_dbl(
        .x = seq_len(n_samples),
        .f = ~ .getPrior(
          i = .x,
          sampler_x0 = sampler_x0,
          sampler_x1 = sampler_x1
        )
      )
    } else{
      SATE_prior <-
        purrr::map_dbl(
          .x = seq_len(n_samples),
          .f = ~ .getPrior(
            i = .x,
            sampler_x0 = sampler_x0,
            sampler_x1 = sampler_x1
          )
        )
    }
    return(SATE_prior)
  }


SATE_prior <-
  getTreatmentPrior(
    dbarts_data = train_data,
    treatment = "treatment",
    n_samples = 10000,
    parallel = TRUE
  )


ggplot(data = tibble(value = SATE_prior / sd(my_data$Y)), aes(x = value)) +
  geom_histogram() +
  geom_vline(xintercept = mean(my_data$tau) / sd(my_data$Y),
             linetype = "dashed") +
  ggtitle("Prior Distribution of the Sample Average Treatment Effect") +
  xlab("Standard Deviations")

```

