---
title: "Working with dbarts Saved Trees"
author: "Vincent Dorie"
date: "01/23/2022"
output: pdf_document
bibliography: references.bib
urlcolor: blue
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Working with dbarts Saved Trees}
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

When a `dbarts` model is fit with the `keepTrees` option set to `TRUE`, the sampler will return those trees in a "flat" format that corresponds to a depth-first, left-hand traversal. Recursion can be used to map functions the nodes in this structure, aggregating statistics of interest.

## Simulation

To illustrate, we generate a fake dataset according to the "Friedman 1" model [@frie:1991].
```{r simulateData}
f <- function(x)
    10 * sin(pi * x[,1] * x[,2]) + 20 * (x[,3] - 0.5)^2 + 10 * x[,4] + 5 * x[,5]

set.seed(99)
sigma <- 1.0
n     <- 100

x <- matrix(runif(n * 10), n, 10)
y <- rnorm(n, f(x), sigma)

data <- data.frame(x, y)
```

## Model Fitting

In order to interrogate the trees, they must be saved when the model is fit. This is accomplished by setting:

* For `bart`: `keeptrees = TRUE`
* For `bart2`: `keepTrees = TRUE`
* For a custom `dbartsSampler`, `control = dbartsControl(keepTrees = TRUE)`

In the context of our fake data, with small sample numbers for illustrative purposes, a model can be fit thusly:
```{r fitModel}
library(dbarts, quietly = TRUE)

bart.1 <- bart(y ~ ., data,
               ndpost = 4,   # number of posterior samples
               nskip = 1000, # number of "warmup" samples to discard
               nchain = 2,   # number of independent, parallel chains
               nthread = 1,  # units of parallel execution
               ntree = 3,    # number of trees per chain
               seed = 2,     # chosen to generate a deep tree
               keeptrees = TRUE,
               verbose = FALSE)
```

## Extracting Trees

The extract function accepts as a `type` the value `"trees"`. If present, the arguments `chainNums`, `sampleNums`, and/or `treeNum`s can be used to extract only a subset of trees.
```{r extractTrees}
trees <- extract(bart.1, "trees")
```

## Flattened Trees

The `trees` data frame corresponds to a depth-first, left-hand-side tree traversal.
```{r printFlattenedTrees}
print(head(trees, n = 10))
```

The columns refer to:

* `chain`, `sample`, `tree` - index variables
* `n` - number of observations in node
* `var` - either the index of the variable used for splitting or -1 if the node is a leaf
* `value` - either the value such that observations less than or equal to it are sent down the left path of the tree or the predicted value for a leaf node

The mapping between the values of `var` and the variable names can be looked up in the internal copy of the data that the sampler stores. This can be found in a fitted model as the element `fit$data@x`, as seen below.

## Tree Traversal

A useful technique for processing trees is [recursion](https://en.wikipedia.org/wiki/Recursion_%28computer_science%29). By having the function return the number of nodes it has processed, it is possible to advance from the left-hand-side to the right by skipping ahead the appropriate number of rows. For example:

```{r rebuildTrees}
# Turns a flatted tree data frame into a list of lists, or a "natural" tree
# structure.
rebuildTree <- function(tree, object) {
    # Define a worker function that will be recursively called on every node.
    rebuildTreeRecurse <- function(tree) {
        node <- list(
            value = tree$value[1],
            n     = tree$n[1]
        )
        # Check node if is a leaf, and if so return early.
        if (tree$var[1] == -1) {
            node$n_nodes <- 1
            return(node)
        }
        
        node$var <- variableNames[tree$var[1]]
        
        # By removing the current row, we can recurse down the left branch.
        headOfLeftBranch <- tree[-1,]
        left <- rebuildTreeRecurse(headOfLeftBranch)
        n_nodes.left <- left$n_nodes
        left$n_nodes <- NULL
        node$left <- left
        
        # The right branch is obtained by advancing past the left nodes.
        headOfRightBranch <- tree[seq.int(2 + n_nodes.left, nrow(tree)),]
        right <- rebuildTreeRecurse(headOfRightBranch)
        n_nodes.right <- right$n_nodes
        right$n_nodes <- NULL
        node$right <- right
        
        node$n_nodes <- 1L + n_nodes.left + n_nodes.right
        
        return(node)
    }
    variableNames <- colnames(object$fit$data@x)
    
    result <- rebuildTreeRecurse(tree)
    result$n_nodes <- NULL
    return(result)
}

treeOfInterest <- subset(trees, chain == 1 & sample == 3 & tree == 1)
print(rebuildTree(treeOfInterest, bart.1))
```

Using a `by` statement, it is possible to "rebuild" all trees at once:
```{r rebuildAllTrees}
allTrees <- by(
    data    = trees,
    INDICES = trees[,c("chain", "sample", "tree")],
    FUN     = rebuildTree, 
    object  = bart.1
)
# One way to index the result of this:
#    allTrees[chain = "1", sample = "2", tree = "3"]
```

## Plotting Trees

`dbartsSampler` objects have a `plotTree` method that can be used to visualize single trees:

```{r plotTree}
bart.1$fit$plotTree(chainNum = 1, sampleNum = 3, treeNum = 1)
```

## Advanced Traversal

The following function can be used to map an arbitrary function over a tree.

```{r mapOverNodes}
mapOverNodes <- function(tree, f, ...) {
    mapOverNodesRecurse <- function(tree, depth, f, ...) {
        node <- list(
            value = tree$value[1],
            n = tree$n[1],
            depth = depth
        )
        if (tree$var[1] == -1) {
            node$n_nodes <- 1
            node$f.x <- f(node, ...)
            return(node)
        }
        node$var <- tree$var[1]
        node$f.x <- f(node, ...)
        
        headOfLeftBranch <- tree[-1,]
        left <- mapOverNodesRecurse(headOfLeftBranch, depth + 1, f, ...)
        n_nodes.left <- left$n_nodes
        left$n_nodes <- NULL
        node$left <- left

        
        headOfRightBranch <- tree[seq.int(2 + n_nodes.left, nrow(tree)),]
        right <- mapOverNodesRecurse(headOfRightBranch, depth + 1, f, ...)
        n_nodes.right <- right$n_nodes
        right$n_nodes <- NULL
        node$right <- right
        
        node$n_nodes <- 1 + n_nodes.left + n_nodes.right
        return(node)
    }
    result <- mapOverNodesRecurse(tree, 1, f, ...)
    result$n_nodes <- NULL
    return(result)
}
```

As an example of its usage, the following function aggregates all ancestor/descendant relationships in a tree. In a data object - here an R environment - it keeps track of the current state of traversal. This includes the variables that have been used for splits above the current node and also includes the current node's depth, which is used to detect backtracking.
```{r observeInteractions}
observeInteractions <- function(node, ...) {
    if (is.null(node$var)) return(NULL)

    interactionData <- list(...)$interactionData
    # Make the current node visibile inside the environment.
    interactionData$node <- node
    with(interactionData, {
        if (node$depth <= currentDepth) {
            # If true, we have backtracked to go down the right branch, so we
            # remove the variables from the left branch.
            currentVariables <- currentVariables[seq_len(node$depth - 1)]
        }
        if (length(interactionData$currentVariables) > 0) {
            # This is a brute-force way of updating the following indices,
            # relying on the column-major storage order that R uses:
            #     hasInteraction[currentVariables,,drop = FALSE][,node$var]
            updateIndices <- currentVariables +
                (node$var - 1) * nrow(hasInteraction)
            hasInteraction[updateIndices] <- TRUE
        }
        currentVariables <- c(currentVariables, node$var)
        currentDepth <- node$depth
    })
    rm("node", envir = interactionData)
    
    # Since the function is used for its side effects, there isn't a return
    # value.
    return(NULL)
}

numVariables  <- ncol(bart.1$fit$data@x)
variableNames <- colnames(bart.1$fit$data@x)

# Define this as an environment as they are mutable
interactionData <- list2env(list(
    currentDepth = 0,
    currentVariables = integer(),
    hasInteraction = matrix(
        data = FALSE,
        ncol = numVariables, nrow = numVariables,
        dimnames = list(ancestor = variableNames, descendant = variableNames)
    )
))

invisible(mapOverNodes(
    treeOfInterest,
    observeInteractions,
    interactionData = interactionData
))
```

After execution, the boolean matrix in the interaction data environment will contain all ancestor/descendant relationships in this tree.

```{r printObserveInteractionResults}
print(interactionData$hasInteraction)
```

Finally, the entire process can be wrapped in a `by` statement and the results aggregated across all trees in order to count the number of times variables have specific relationships.

## References
